Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-11 examples.tex
+

\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
%$\assert{\ulcorner \mathsf{aligned maddr} \urcorner \ast \mathsf{r14} \mapsto_{\textsf{r}} \textsf{r14v} \ast \mathsf{r13} \mapsto_{\textsf{r}} \textsf{maddr} \ast \mathsf{rdi} \mapsto_{\textsf{r}} \textsf{rdiv} \ast \mathsf{rax} \mapsto_{\textsf{r}} \textsf{raxv}}$
%$\assert{\ulcorner \ptablestore !! \textsf{maddr} = \textsf{None} \urcorner \ast \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \textsf{Pf}} $
\section{Experiment}
\label{sec:experiment}
To both validate and demonstrate the value of the modal approach to reasoning about virtual memory management, we study several distillations of real concerns of virtual memory managers.
Recall from Section \ref{sec:logic} that virtual points-to assertions work just like regular points-to assertions, by design.

\begin{comment}
\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
This still works with the modality, since that invariant is also semantically a predicate on a page table root.
Let's call this v2.
\todo[inline]{the current course of action is to finish the WPs for v1 and try an example that should work with either approach (like the task switch). This is a good basic sanity check on our approaches. Once that's done, then we work out the details of v2, which should require \emph{refactoring} the same sorts of reasoning we had for v1, rather than totally different proofs from v1.}
\end{comment}

@@

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred)  \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \texttt{maddr} = \texttt{None}\urcorner}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast }_{\rtv}$

>>>>

%$\assert{\ulcorner \mathsf{aligned maddr} \urcorner \ast \mathsf{r14} \mapsto_{\textsf{r}} \textsf{r14v} \ast \mathsf{r13} \mapsto_{\textsf{r}} \textsf{maddr} \ast \mathsf{rdi} \mapsto_{\textsf{r}} \textsf{rdiv} \ast \mathsf{rax} \mapsto_{\textsf{r}} \textsf{raxv}}$
%$\assert{\ulcorner \ptablestore !! \textsf{maddr} = \textsf{None} \urcorner \ast \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \textsf{Pf}} $
\section{Experiment}
\label{sec:experiment}
To both validate and demonstrate the value of the modal approach to reasoning about virtual memory management, we study several distillations of real concerns of virtual memory managers.
Recall from Section \ref{sec:logic} that virtual points-to assertions work just like regular points-to assertions, by design.

%\begin{comment}
%\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
%\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
%\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
%We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
%This still works with the modality, since that invariant is also semantically a predicate on a page table root.
%Let's call this v2.
%\end{comment}

@@

\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}

@@

\{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred)  \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \texttt{maddr} = \texttt{None}\urcorner\}_{\rtv}

@@

$\{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast \}_{\rtv}$

### 2023-04-11 syntaxsemantics.tex
section 3
+

 Definition translate (m: $\sigma.\mathcal{M}$) (root: word 64) (w: word 64): address :=
      let plm4e := translate_top_level m root  w in
      let pdpte := translate_from_l4e m pml4e w in 
      let pde := translate_from_l3e m pdpte w in
      let pte: = translate_from_l2e m pde w in
      let addr := translate_from_l1e m pte w in

@@

we see translation physical memory $\mathsf{load}$ for each level of page-table-walk, which is basically a table lookup with

@@

   Definition load (mem: $\sigma.\mathcal{M}$) (w: { w: word 64 | aligned w }):word 64 + MemFail :=
    match  $\sigma$.mem !! (pfn (proj1_sig w)) with

@@

  Definition store (mem: $\sigma.\mathcal{M}$) (w: { w: word 64 | aligned w }) (v: word 64):mem64 * (unit + MemFail) :=
    match $\sigma$.mem !! (pfn (proj1_sig w)) mem with

@@

        (($\sigma.\mathcal{M}$<[(pfn (proj1_sig w)) := (inl (fun off =>
           if (word_dec off (page_off (proj1_sig w))) then v else zero 64))]> mem), inl tt)

@@

        (($\sigma.\mathcal{M}$<[(pfn (proj1_sig w)) := (inl (fun off =>
           if (word_dec off (page_off (proj1_sig w))) then v else zero 64))]> mem), inl tt)
      | Some (inr fault) => (mem, inr fault)

@@

  \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad \qquad\qquad\qquad 

@@

  \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad \qquad\qquad\qquad 

@@

  Register instructions, on the other hand, are implemented as $\sigma.\mathcal{R}$ lookup and store
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToRegFromReg})} \qquad \qquad\qquad\qquad 
  \textsf{mov ireg64 ireg64} \\
  \hbox{(\TirNameStyle{WriteToCtlRegFromReg})} \qquad \qquad\qquad\qquad 
  \textsf{mov creg64 ireg64}
    \hbox{(\TirNameStyle{WriteToRegFromCtlReg})} \qquad \qquad\qquad\qquad 
  \textsf{mov ireg64 creg64}
  \end{array}
\]
for different types of registers (e.g. control and general).
%
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}

>>>>

 Definition translate (root: word 64) (w: word 64): address :=
      let plm4e := translate_top_level $\sigma.\mathcal{M}$ root  w in
      let pdpte := translate_from_l4e $\sigma.\mathcal{M}$ pml4e w in 
      let pde := translate_from_l3e $\sigma.\mathcal{M}$ pdpte w in
      let pte: = translate_from_l2e $\sigma.\mathcal{M}$ pde w in
      let addr := translate_from_l1e $\sigma.\mathcal{M}$ pte w in

@@

we see physical memory $\mathsf{load}$ for each level of page-table-walk concluded with alignment of the returned address.

@@

   Definition load (w: { w: word 64 | aligned w }):word 64 + MemFail :=
    match  $\sigma.\mathcal{M}$ !! (pfn (proj1_sig w)) with

@@

  Definition store (w: { w: word 64 | aligned w }) (v: word 64):mem64 * (unit + MemFail) :=
    match $\sigma.\mathcal{M}$ !! (pfn (proj1_sig w)) with

@@

        ((<[(pfn (proj1_sig w)) := (inl (fun off => if (word_dec off (page_off (proj1_sig w)))
                                        then v else zero 64))]> $\sigma.\mathcal{M}$), inl tt)

@@

        ((<[(pfn (proj1_sig w)) := (inl (fun off => if (word_dec off (page_off (proj1_sig w)))
                                        then v else zero 64))]> $\sigma.\mathcal{M}$), inl tt)
      | Some (inr fault) => ($\sigma.\mathcal{M}$, inr fault)

@@

  \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad

@@

  \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 

@@

  Register instructions, on the other hand, are implemented as $\sigma.\mathcal{R}$ lookup and store
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToRegFromReg})} \qquad 
  \textsf{mov ireg64 ireg64} \\
  \hbox{(\TirNameStyle{WriteToCtlRegFromReg})} \qquad 
  \textsf{mov creg64 ireg64}\\
    \hbox{(\TirNameStyle{WriteToRegFromCtlReg})} \qquad 
  \textsf{mov ireg64 creg64}
  \end{array}
\]
for different types of registers (e.g. control and general).
%
%\subsection{Machine State under Address Translation}
%\label{sec:selectedinstrsemantics}
%Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
%\begin{itemize}
%\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
%\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
%\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
%\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
%\end{itemize}
